---
title: "MDTS4313_737_4"
author: "SULAGNA ROY"
date: "2026-08-16"
output: pdf_document
---
# Problem Set 4 : Kernel Smoother
# ------------------------------------
## 1. Write the following lines of code in R.
##      set . seed (1234)
##      X=runif(100 ,5 ,15)
##      Y=5*sin(X)+23*cos(X)^2+rnorm(100 ,0 ,5)
##      Sim.1=data.frame(X=X,Y=Y)
##      reg=function(x){5*sin(x)+23*cos(x)^2}
##  The function “reg” represents the truth (i.e., the regression function) of
##  the simulated dataset.

### Seed for Reproducible Outputs
```{r}
set.seed (1234) 
```

### 100 obs.s on predictor X from U(5,15)
```{r}
X=runif(100,5,15)
```

### 100 obs. on response Y (errors~N(mean=0,sd=5))
```{r}
Y=(5*sin(X))+(23*cos(X)^2)+rnorm(100,0,5) 
```

### Full Data Set
```{r}
Sim.1=data.frame(X=X,Y=Y)
```


# 100 obs. on the true regression function.
```{r}
reg=function(x){5*sin(x)+23*cos(x)^2}
```

# ------------------------------------

## 2. Divide the X values into parts. Call the first 80 values the training test.
##    Denote them by X.train. The rest of the 20 values constitute the test set.
##    Call them X.test.

# Training Set of Predictor X
```{r}
X.train=X[1:80]
```

# Test Set of Predictor X
```{r}
X.test=X[81:100]
```

# ------------------------------------

## 3. Divide the Y values in training set (Y.train) and test set (Y.test) in the
##    similar way.

# Training Set of Response Y
```{r}
Y.train=Y[1:80]
```

# Test Set of Response Y
```{r}
Y.test=Y[81:100]
```

# ------------------------------------

## 4. Plot the data such that the points corresponding to the training set and
##    the test set are represented with different symbols.
```{r}
# Filled circle for training 
plot(X.train,Y.train,pch = 16,col="blue",
     xlim = range(X),ylim=range(Y),
     xlab="X",ylab="Y",main="Training and Test Data") 

# Add test observations(# Triangle for test ) 
points(X.test,Y.test,pch=17,col="red") 

# Add legend 
legend("topright",legend=c("Training Data","Test Data"),pch = c(16, 17),col = c("blue", "red"),cex=0.8)

```

# ------------------------------------

## 5. Plot the regression function in the same graph.
```{r}
# Create a sequence of X values for a smooth regression curve 
x.grid=seq(min(X),max(X),length.out=500) 

# Calculate the true regression function 
y.true=reg(x.grid)

# Filled circle for training 
plot(X.train,Y.train,pch = 16,col="blue",cex=0.7,
     xlim = range(X),ylim=range(Y),
     xlab="X",ylab="Y",main="Training and Test Data") 

# Add test observations(# Triangle for test ) 
points(X.test,Y.test,pch=17,col="red",cex=0.7) 

# Add true regression function 
lines(x.grid,y.true,col="black",lwd = 2) 

# Add legend 
legend("topleft",
       legend=c("Training Data","Test Data","True Regression Function"), 
       pch=c(16,17,NA), 
       lty=c(NA,NA,1), 
       col=c("blue","red","black"), 
       lwd=c(NA,NA,2),
       cex=0.5)
```

# ------------------------------------

## 6. Apply the Kernel smoother algorithm on the training dataset with the 
##    following kernels
##    (a) Gaussian with sigma = 0.1,0.25,0.5,1,2.
##    (b) Uniform Kernel with h = 0.1,0.25,0.5,1,2 .
##    (c) Triangular Kernel with h = 0.1,0.25,0.5,1,2.
##    (d) Epanechnikov Kernel with h = 0.1,0.25,0.5,1,2.
##    For each case, plot the estimates in separate graphs. Comment on your
##    findings (i.e., how do the estimates change with changing h(sigma)?)

# 6.1 — Define Kernel Smoother
```{r}
# Kernel Smoother Function

kernel.smooth <- function(x, y, x0, h, kernel) {
  
  u <- (x0 - x) / h
  
  if(kernel == "Gaussian")
    w <- exp(-u^2 / 2)
  
  if(kernel == "Uniform")
    w <- ifelse(abs(u) <= 1, 1, 0)
  
  if(kernel == "Triangular")
    w <- ifelse(abs(u) <= 1, 1 - abs(u), 0)
  
  if(kernel == "Epanechnikov")
    w <- ifelse(abs(u) <= 1, 1 - u^2, 0)
  
  sum(w * y) / sum(w)
}
```

# 6.2 — Set bandwidths and prediction grid
```{r}
h.val <- c(0.1, 0.25, 0.5, 1, 2)

x.grid <- seq(min(X.train), max(X.train), length.out = 200)
```

# 6.3 — Gaussian Kernel
```{r}
gauss.est <- matrix(NA, nrow = length(x.grid), ncol = 5)

for(j in 1:5)
  gauss.est[,j] <- sapply(x.grid, function(x)
    kernel.smooth(X.train, Y.train, x, h.val[j], "Gaussian"))

matplot(x.grid, gauss.est, type = "l", lty = 1,
        xlab = "X", ylab = "Y",
        main = "Gaussian Kernel")

legend("topright", legend = paste(expression(sigma), h.val),
       lty = 1, col = 1:5)
```
# 6.4 — Uniform Kernel
```{r}
uniform.est <- matrix(NA, nrow = length(x.grid), ncol = 5)

for(j in 1:5)
  uniform.est[,j] <- sapply(x.grid, function(x)
    kernel.smooth(X.train, Y.train, x, h.val[j], "Uniform"))

matplot(x.grid, uniform.est, type = "l", lty = 1,
        xlab = "X", ylab = "Y",
        main = "Uniform Kernel")

legend("topright", legend = paste("h =", h.val),
       lty = 1, col = 1:5)
```
# 6.5 — Triangular Kernel
```{r}
tri.est <- matrix(NA, nrow = length(x.grid), ncol = 5)

for(j in 1:5)
  tri.est[,j] <- sapply(x.grid, function(x)
    kernel.smooth(X.train, Y.train, x, h.val[j], "Triangular"))

matplot(x.grid, tri.est, type = "l", lty = 1,
        xlab = "X", ylab = "Y",
        main = "Triangular Kernel")

legend("topright", legend = paste("h =", h.val),
       lty = 1, col = 1:5)
```
# 6.6 — Epanechnikov Kernel
```{r}
epa.est <- matrix(NA, nrow = length(x.grid), ncol = 5)

for(j in 1:5)
  epa.est[,j] <- sapply(x.grid, function(x)
    kernel.smooth(X.train, Y.train, x, h.val[j], "Epanechnikov"))

matplot(x.grid, epa.est, type = "l", lty = 1,
        xlab = "X", ylab = "Y",
        main = "Epanechnikov Kernel")

legend("topright", legend = paste("h =", h.val),
       lty = 1, col = 1:5)
```

## Comment:

For smaller values of h (or $\sigma$), the kernel smoother follows the observations more closely and produces a more variable or wiggly estimate. As h increases, more observations receive substantial weight, resulting in a smoother curve. Very large bandwidths may oversmooth the data and fail to capture important features of the regression function.

# ------------------------------------

## 7. For each case, compute the training and test errors.

# 7.1 — Error Function
```{r}
MSE <- function(actual, predicted)
  mean((actual - predicted)^2)
```

# 7.2 — Calculate errors for all kernels
```{r}
errors <- data.frame()

for(k in c("Gaussian", "Uniform", "Triangular", "Epanechnikov")) {
  
  for(h in h.val) {
    
    train.pred <- sapply(X.train, function(x)
      kernel.smooth(X.train, Y.train, x, h, k))
    
    test.pred <- sapply(X.test, function(x)
      kernel.smooth(X.train, Y.train, x, h, k))
    
    errors <- rbind(errors,
                    data.frame(Kernel = k,
                               h = h,
                               Training.Error = MSE(Y.train, train.pred),
                               Test.Error = MSE(Y.test, test.pred)))
  }
}

errors
```

# 7.3 — Find best bandwidth for each kernel
```{r}
errors[order(errors$Test.Error), ]
```

# 7.4 - the best value for each kernel
```{r}
do.call(rbind,
        lapply(split(errors, errors$Kernel),
               function(z) z[which.min(z$Test.Error), ]))
```

## Comment :

Smaller bandwidths give lower training errors but may overfit the data. Increasing the bandwidth produces smoother estimates and may reduce test error up to an optimal value. Very large bandwidths can cause underfitting and increase the test error. Thus, the bandwidth with minimum test error is preferred.

# ------------------------------------

## 8.  Use the ksmooth function in R to compare the results with your code for Gaussian Kernel only. 
##     (Home Work: try to find any other inbuilt function in R to implement the other kernels.)

# 8.1 — Gaussian using ksmooth()
```{r}
ks <- list()

for(h in h.val) {
  
  ks[[as.character(h)]] <-
    ksmooth(X.train, Y.train,
            kernel = "normal",
            bandwidth = h,
            x.points = x.grid)
}
```

# 8.2 — Plot ksmooth() results
```{r}
plot(X.train, Y.train,
     pch = 16,cex=0.5,
     xlab = "X", ylab = "Y",
     main = "Gaussian Kernel using ksmooth()")

for(j in 1:5)
  lines(ks[[j]]$x, ks[[j]]$y,
        lty = 1,lwd=1.2,
        col=j)

legend("topright",
       legend = paste(expression(sigma), h.val),
       lty = 1,
       col=1:5,
       cex=0.5)
```

```{r}
ks05 <- ksmooth(X.train, Y.train,
                kernel = "normal",
                bandwidth = 0.5,
                x.points = x.grid)

plot(x.grid, gauss.est[,3],
     type = "l",ylim=c(-3.5,40),
     xlab = "X", ylab = "Y",
     main = "Comparison: Gaussian Kernel")

lines(ks05$x, ks05$y, lty = 2)

legend("topright",
       legend = c("My Kernel", "ksmooth"),
       lty = 1:2)
```

## Comment :

The Gaussian kernel smoother obtained using our own kernel-smoothing function is very similar to the result obtained using the built-in ksmooth() function in R. As the bandwidth $/sigma$ increases, both methods produce smoother curves, while smaller values of $/sigma$ give more flexible and wiggly estimates. Thus, ksmooth() provides results consistent with the manually implemented Gaussian kernel smoother. The question specifically asks for this comparison for the Gaussian kernel.

# ------------------------------------

## 9.   Repeat the procedure 50 times. Each time, the training set and the test set should be chosen randomly 
##      from the given sample. Find the average of the training errors and 
##      the testing errors thus obtained for each case and comment.

# 9.1 — 50 repetitions
```{r}
set.seed(1234)

all.errors <- data.frame()

for(rep in 1:50) {
  
  ind <- sample(1:100, 80)
  
  x.tr <- X[ind]
  y.tr <- Y[ind]
  x.te <- X[-ind]
  y.te <- Y[-ind]
  
  for(k in c("Gaussian", "Uniform",
             "Triangular", "Epanechnikov")) {
    
    for(h in h.val) {
      
      train.pred <- sapply(x.tr, function(x)
        kernel.smooth(x.tr, y.tr, x, h, k))
      
      test.pred <- sapply(x.te, function(x)
        kernel.smooth(x.tr, y.tr, x, h, k))
      
      all.errors <- rbind(all.errors,
                          data.frame(Rep = rep,
                                     Kernel = k,
                                     h = h,
                                     Train.Error =
                                       MSE(y.tr, train.pred),
                                     Test.Error =
                                       MSE(y.te, test.pred)))
    }
  }
}
```

# 9.2 — Average errors
```{r}
avg.errors <- aggregate(cbind(Train.Error, Test.Error) ~ Kernel + h,
                        data = all.errors,
                        FUN = mean)

avg.errors
```

# 9.3 — Find best bandwidth for each kernel
```{r}
do.call(rbind,
        lapply(split(avg.errors, avg.errors$Kernel),
               function(z) z[which.min(z$Test.Error), ]))
```

## Comment :

After repeating the random training-test split 50 times, the average errors provide a more reliable comparison of the kernels. Small bandwidths generally produce lower training error because the smoother follows the observations closely, but they may produce larger test error due to overfitting. Increasing the bandwidth smooths the estimates and can reduce test error up to an optimal bandwidth. Beyond this point, excessive smoothing can increase both bias and test error.

# --------------------------------------------------------

## 10.  Consider the “Boston” dataset in the “MASS” library in R and let the
##      variable “medv” be the response (Y ) and the variable “lstat” be the predictor (X). 
##      Use 22-fold cross-validation method to find the best choice of $/sigma$ for applying the Gaussian Kernel smoother.

# 10.1 — Load Boston data
```{r}
library(MASS)

data(Boston)

X.b <- Boston$lstat
Y.b <- Boston$medv
```

# 10.2 — 22-fold cross-validation
```{r}
set.seed(1234)

K <- 22
fold <- sample(rep(1:K, length.out = nrow(Boston)))

sigma.val <- c(0.1, 0.25, 0.5, 1, 2)

cv.error <- numeric(length(sigma.val))

for(j in 1:length(sigma.val)) {
  
  err <- numeric(K)
  
  for(k in 1:K) {
    
    train <- which(fold != k)
    test <- which(fold == k)
    
    pred <- sapply(X.b[test], function(x)
      kernel.smooth(X.b[train],
                    Y.b[train],
                    x,
                    sigma.val[j],
                    "Gaussian"))
    
    err[k] <- mean((Y.b[test] - pred)^2)
  }
  
  cv.error[j] <- mean(err)
}

cv.result <- data.frame(sigma = sigma.val,
                        CV.Error = cv.error)

cv.result
```
# 10.3 — Find best sigma
```{r}
cv.result[which.min(cv.result$CV.Error), ]
```
# 10.4 — Plot CV Error
```{r}
plot(sigma.val, cv.error,
     type = "b",
     pch = 16,
     xlab = expression(sigma),
     ylab = "22-Fold CV Error",
     main = "Choosing sigma using Cross-Validation")
```

```{r}
cv.result[which.min(cv.result$CV.Error), ]
```

## Comment :

The optimal value of sigma is selected as the value corresponding to the minimum 22-fold cross-validation error. This value provides the best balance between under-smoothing and over-smoothing for predicting medv from lstat.

# --------------------------------------------------------